---
title: "Phosphoproteomics"
output: html_notebook

---
***
#### Download raw data from Uppmax
```{bash}
rsync -r noahe@rackham.uppmax.uu.se:/proj/g2020004/private/student_projects/phosphoproteomics /home/noah/phosphoproteomics_project/
```

#### Setting working directory and libraries
```{r}
setwd("~/phosphoproteomics_project")
RequiredPackages <- c() #yet empty
for (i in RequiredPackages) { #Installs packages if not yet installed
    if (!require(i, character.only = TRUE)) install.packages(i)
}
```

#### Load raw data and change default separator to ";"
```{r}
raw_data_phospho <- read.csv(file = "/home/noah/phosphoproteomics_project/phosphoproteomics/Dataset_Phospho.csv", sep = ";")
raw_data_proteom <- read.csv(file = "/home/noah/phosphoproteomics_project/phosphoproteomics/Dataset_Proteome.csv", sep = ";")
```
#### If loop to omit na.value from all Abundance columns. 
- grep line selects all columns that contain the string "Abundance" and saves the column number in *Abundance_columns*
- The for loop then selects all row were there is *no* NA value in one of the Abundance columns and copies them to data_phospho to leave the raw data untouched.
- **!(is.na(raw_data_phospho[,i]))** gives out all the rows that don't have an NA value, these get selected
```{r}
Abundance_columns <- grep("Abundance", colnames(raw_data_phospho) )
for (i in Abundance_columns) {
  data_phospho <- raw_data_phospho[!(is.na(raw_data_phospho[,i])), ]
  }
```

#### Calculate Mean for Abundance and adding it as column and Ratio LIF/Control
- grep again Abundance but this time the Control and LIF (case) seperatly. 
- calculate row means each and the ration between the means.

```{r}
Abundance_columns <- grep("Abundance.Control", colnames(data_phospho))
data_phospho$Abudance.Mean.Control <- rowMeans(data_phospho[,c(Abundance_columns)])
Abundance_columns <- grep("Abundance.LIF", colnames(data_phospho))
data_phospho$Abudance.Mean.LIF <- rowMeans(data_phospho[,c(Abundance_columns)])
data_phospho$Abudance.Ratio <- data_phospho$Abudance.Mean.LIF / data_phospho$Abudance.Mean.Control
```

#### Calculate p-values
- *t.test* to compare all 3 Control <-> 3 Case values. If they are different then the pvalue < 0.05 **->** indicating a real differences between case & control. If p-value is big the difference is by chance😄 I tried it for row 188 which has a Ration nearly equal to 1 (1.02...) indicating no difference in case vs. control and the p value ~ 0.7 (really bad) also indicating no difference.
```{r}
hist(data_phospho$Abudance.Ratio)
i = 1
for (row in 1:nrow(data_phospho)) {
  p_value <- t.test(data_phospho[i,10:12], data_phospho[i,13:15]) 
  #tests if mean of control is different from mean of LIF (case) and assigns it to list p_value
  data_phospho$Abudance.pvalue[i] <- p_value$p.value 
  #selects p.value from list and assigns it to new column"Abudance.pvalue" in the correct row
  i = 1+i
}
```
#### Subsetting data_phospho for ratio values >= 2
- a. Remove peptides without quantitative values -> Already done in "If loop to omit na.value from all Abundance columns" thereby creating the data_phospho
```{r}
more_phosphorylated <- subset.data.frame(data_phospho, subset = data_phospho$Abudance.Ratio >= 2)
```

