---
title: "Phosphoproteomics"
output: html_notebook
---
Download raw data
```{bash}
rsync -r noahe@rackham.uppmax.uu.se:/proj/g2020004/private/student_projects/phosphoproteomics /home/noah/phosphoproteomics_project/
```

Setting working directory and libraries
```{r}
setwd("~/phosphoproteomics_project")
RequiredPackages <- c() #yet empty
for (i in RequiredPackages) { #Installs packages if not yet installed
    if (!require(i, character.only = TRUE)) install.packages(i)
}
```

Load raw data and change default separator to ";"
```{r}
raw_data_phospho <- read.csv(file = "/home/noah/phosphoproteomics_project/phosphoproteomics/Dataset_Phospho.csv", sep = ";")
raw_data_proteom <- read.csv(file = "/home/noah/phosphoproteomics_project/phosphoproteomics/Dataset_Proteome.csv", sep = ";")
```
If loop to omit na.value from all Abundance columns
```{r}
Abundance_columns <- grep("Abundance", colnames(raw_data_phospho) )
for (i in Abundance_columns) {
  data_phospho <- raw_data_phospho[!(is.na(raw_data_phospho[,i])), ]
  }
```

Calculate Mean for Abundance and adding it as column and Ratio LIF/Control
```{r}
Abundance_columns <- grep("Abundance.Control", colnames(data_phospho))
data_phospho$Abudance.Mean.Control <- rowMeans(data_phospho[,c(Abundance_columns)])
Abundance_columns <- grep("Abundance.LIF", colnames(data_phospho))
data_phospho$Abudance.Mean.LIF <- rowMeans(data_phospho[,c(Abundance_columns)])
data_phospho$Abudance.Ratio <- data_phospho$Abudance.Mean.LIF / data_phospho$Abudance.Mean.Control
```

Calculate p-values
```{r}
hist(data_phospho$Abudance.Ratio)

```



Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
