---
title: "Phosphoproteomics"
output: html_notebook
---

------------------------------------------------------------------------

#### Download raw data from Uppmax

```{bash}
rsync -r noahe@rackham.uppmax.uu.se:/proj/g2020004/private/student_projects/phosphoproteomics /home/noah/phosphoproteomics_project/
```

------------------------------------------------------------------------

#### Setting working directory and libraries

```{r, warning=FALSE, results='hide'}
setwd("~/phosphoproteomics_project")
RequiredPackages <- c("BiocManager", "dplyr", "igraph", "tidyr","ggplot2","ggrepel") 
#Installs packages if not yet installed
for (i in RequiredPackages) { 
    if (!require(i, character.only = TRUE)) 
    install.packages(i)
}
BiocManager::install(version = "3.15")
library(BiocManager)
BiocManager::install("STRINGdb")
library(tidyr)
library(dplyr)
library(igraph)
library(STRINGdb)
library(ggplot2)
library(ggrepel)
remove(RequiredPackages)
```

------------------------------------------------------------------------

#### Load raw data and change default separator to ";"

```{r}
raw_data_phospho <- read.csv(file = "/home/noah/phosphoproteomics_project/phosphoproteomics/Dataset_Phospho.csv", sep = ";")
raw_data_proteom <- read.csv(file = "/home/noah/phosphoproteomics_project/phosphoproteomics/Dataset_Proteome.csv", sep = ";")
```

------------------------------------------------------------------------

#### If loop to omit na.value from all Abundance columns

-   grep line selects all columns that contain the string "Abundance" and saves the column number in *Abundance_columns*
-   The for loop then selects all row were there is *no* NA value in one of the Abundance columns and copies them to data_phospho to leave the raw data untouched.
-   **!(is.na(raw_data_phospho[,i]))** gives out all the rows that don't have an NA value, these get selected

```{r}
Abundance_columns <- grep("Abundance", colnames(raw_data_phospho) )
for (i in Abundance_columns) {
  data_phospho <- raw_data_phospho[!(is.na(raw_data_phospho[,i])), ]
  }
```

------------------------------------------------------------------------

#### Calculate Mean for Abundance and adding it as column and Ratio LIF/Control

-   grep again Abundance but this time the Control and LIF (case) seperatly.
-   calculate row means each and the ration between the means.

```{r}
Abundance_columns <- grep("Abundance.Control", colnames(data_phospho))
data_phospho$Abudance.Mean.Control <- rowMeans(data_phospho[,c(Abundance_columns)])

Abundance_columns <- grep("Abundance.LIF", colnames(data_phospho))
data_phospho$Abudance.Mean.LIF <- rowMeans(data_phospho[,c(Abundance_columns)])

data_phospho$Abudance.Ratio <- data_phospho$Abudance.Mean.LIF / data_phospho$Abudance.Mean.Control

remove(Abundance_columns, i)
```

------------------------------------------------------------------------

#### Calculate p-values

-   *t.test* to compare all 3 Control \<-\> 3 Case values. If they are different then the pvalue \< 0.05 **-\>** indicating a real differences between case & control. If p-value is big the difference is by chance😄 I tried it for row 188 which has a Ration nearly equal to 1 (1.02...) indicating no difference in case vs. control and the p value \~ 0.7 (really bad) also indicating no difference.

```{r warning=FALSE}
start_for <- Sys.time()
for (row in 1:nrow(data_phospho)) {
  p_value <- t.test(data_phospho[row,10:12], data_phospho[row,13:15], alternative = "less") 
  #tests if mean of control is different from mean of LIF (case) and assigns it to list p_value
  data_phospho$Abudance.pvalue[row] <- p_value$p.value 
  #selects p.value from list and assigns it to new column"Abudance.pvalue" in the correct row
  }
end_for <- Sys.time()
end_for - start_for #We wanted to compare calculation time to compare the methods

#Alice quicker approach
df_copy <- dplyr::tibble(data_phospho) # Copy df
start_alice <- Sys.time()
df_copy[] <- lapply(df_copy, function(x) as.numeric(as.character(x)))
df_copy$pValues <- apply(df_copy, 1, function(x) t.test(x[10:12],x[13:15], alternative = "less", paired = F)$p.value)
end_alice <- Sys.time()
end_alice-start_alice #Proves that apply is x4 faster then loop
remove(end_alice,end_for,start_alice,start_for,row,p_value, df_copy)
```

------------------------------------------------------------------------

#### Adding a Volcano plot to explore the initial data

```{r}
### Phospho volcano plot
# Calculate LIF/Control ratio
phospho <- data_phospho
phospho$FC <- data_phospho$Abudance.Mean.LIF / data_phospho$Abudance.Mean.Control
phospho$log2_FC <- log2(phospho$FC)
phospho$diffexpressed <- "NO"
phospho$diffexpressed[phospho$log2_FC > 1 & phospho$Abudance.pvalue < 0.05] <- "UP"
phospho$diffexpressed[phospho$log2_FC < -1 & phospho$Abudance.pvalue < 0.05] <- "DOWN"

phospho$delabel <- NA
phospho$delabel[phospho$diffexpressed != "NO"] <- phospho$Master.Protein.Accessions[phospho$diffexpressed != "NO"]

##Addition to generate "real names" not Uniprot IDs
string_db_volcano <- STRINGdb$new( version="11.0b", species=10090, score_threshold=1, input_directory="")
Mapping_phospho <- string_db_volcano$map(phospho, "delabel", removeUnmappedRows = F)
Mapping_phospho <- string_db_volcano$add_proteins_description(Mapping_phospho) #Add´s protein information to STRINGid´s 
#Volcano plot overview
ggplot(data=Mapping_phospho, aes(x=log2_FC, y=-log10(Abudance.pvalue), col=diffexpressed, label=preferred_name))+
  geom_point()+
  theme_minimal()+
  geom_vline(xintercept=c(-1, 1), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red")

ggsave("~/phosphoproteomics_project/results/Zoomed_out_Volcano.png", width = 1500, height = 1000, units = "px")

#Detail/Zoomed in
ggplot(data=Mapping_phospho, aes(x=log2_FC, y=-log10(Abudance.pvalue), col=diffexpressed, label=preferred_name))+
  geom_point()+
  theme_minimal()+
  geom_vline(xintercept=c(-1, 1), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red") + 
  coord_cartesian(xlim = c(1,3.5), ylim = c(0.5, 6))+
  geom_label_repel()

ggsave("~/phosphoproteomics_project/results/Zoomed_in_Volcano.png", width = 1500, height = 1000, units = "px")

### Proteome volcano plot
# Defining function first
volcano_plot_proteome <- function(proteome){
proteome$log2_FC <- log2(proteome$Abundance.Ratio...LIF.....Control.)
proteome$diffexpressed <- "NO"
proteome$diffexpressed[proteome$log2_FC > 1 & proteome$Abundance.Ratio.P.Value...LIF.....Control. < 0.05] <- "UP"
proteome$diffexpressed[proteome$log2_FC < -1 & proteome$Abundance.Ratio.P.Value...LIF.....Control. < 0.05] <- "DOWN"

ggplot(data=proteome, aes(x=log2_FC, y=-log10(Abundance.Ratio.P.Value...LIF.....Control.), col=diffexpressed))+
  geom_point()+
  theme_minimal()+
  geom_vline(xintercept=c(-1, 1), col="red") +
  geom_hline(yintercept=-log10(0.05), col="red")
}
#Printing the volcano plot
volcano_plot_proteome(raw_data_proteom)
remove(phospho, Mapping_phospho, string_db_volcano)
```

------------------------------------------------------------------------

#### Subsetting data_phospho for ratio values \>= 2 and p-values \<= 0.05

-   

    a.  Remove peptides without quantitative values -\> Already done in "If loop to omit na.value from all Abundance columns" thereby creating the data_phospho

-   Here just the data.frame is subset, selecting only Ratios \>= 2 and p-values \<= 0.05

```{r}
##Filter NA out in proteome, Filter out p.values that are non significant(not adjusted)

df_phospho = data_phospho
df_proteome = raw_data_proteom
  
## Filter PHOSPHO
  more_phospho <- subset.data.frame(df_phospho, subset = df_phospho$Abudance.Ratio >= 2) #Filters out all ratios below 2-fold
  more_phospho <- subset.data.frame(more_phospho, subset = more_phospho$Abudance.pvalue <= 0.05) #Filters out all non-significant p-values
  
  ##Filter PROTEOME
  exclude_proteome <- filter(df_proteome, df_proteome$Abundance.Ratio.P.Value...LIF.....Control. <= 0.05 & df_proteome$Abundance.Ratio...LIF.....Control. >= 1) #Filters out all significant p-values, because significant p-values indicate an 
  #Data that we want to exclude, testing if the right data is selected with the volcano plot
  volcano_plot_proteome(exclude_proteome)
  #Anti_join anti_join() return all rows from x without a match in y. 
  more_proteome <- anti_join(df_proteome, exclude_proteome, by = c("Accession" = "Accession"))
  more_proteome <- filter(more_proteome, more_proteome$Abundance.Ratio...LIF.....Control. <= 2)
  #Checking the if the right data is excluded
  volcano_plot_proteome(more_proteome)
  
  #Inner join works too, but semi join is better becasue the proteome columns are not included into the resulting data frame.
  #more_phosphorylated <- dplyr::inner_join( more_phospho, more_proteome, by = c("Master.Protein.Accessions" = "Accession"))
  
  #Semi-join #semi_join() return all rows from x with a match in y. 
  more_phosphorylated <- dplyr::semi_join(more_phospho, more_proteome, by = c("Master.Protein.Accessions" = "Accession"))
  
remove(df_phospho, df_proteome, more_phospho, more_proteome, exclude_proteome)
```

------------------------------------------------------------------------

#### Building network map of the proteins

```{r}
##Lower score_threshold, allowing more connections
string_db_low <- STRINGdb$new(version="11.0b", species=10090, score_threshold=1, input_directory="") #species = mouse identifier, score threshold = if one interaction this will be mapped, no input directory data stored only temporary
hits_low <- string_db_low$map(more_phosphorylated, "Master.Protein.Accessions", removeUnmappedRows = F) #mapping the UNIPROT Ids from the first column to get STRING identifiers

string_db_low$plot_network(hits_low$STRING_id) #plotting network
invisible(string_db_low$get_png(hits_low$STRING_id, file = "~/phosphoproteomics_project/results/Network_low.png")) #Saving photo

##Higher score_threshold, allowing less connections
string_db_high <- STRINGdb$new(version="11.0b", species=10090, score_threshold=200, input_directory="") #species = mouse identifier, score threshold = if one interaction this will be mapped, no input directory data stored only temporary
hits_high <- string_db_high$map(more_phosphorylated, "Master.Protein.Accessions", removeUnmappedRows = F) #mapping the UNIPROT Ids from the first column to get STRING identifiers

string_db_high$plot_network(hits_high$STRING_id) #plotting network
invisible(string_db_high$get_png(hits_high$STRING_id, file = "~/phosphoproteomics_project/results/Network_high.png")) #Saving photo
```

------------------------------------------------------------------------

#### Way to identify the Proteins Stat5b is connected to.

-   There is also a link to the [STRINGdb website for less stringent threshold = 1](https://version-11-0b.string-db.org/cgi/link?to=2AE339B7E8089CA3) with the Proteins connected to Stat5b in this dataset.
-   There is also a link to the [STRINGdb website for more stringent threshold = 200](https://version-11-0b.string-db.org/cgi/link?to=38BC613D55C2F293) with the Proteins connected to Stat5b in this dataset.

```{r}
#Finding String id identifier for Stat5b and assigning it to the a variable
i <- grep("P42232", hits_low$Master.Protein.Accessions) 
Stat5b <- string_db_low$mp("P42232") #mapping the UNIPROT Ids 

## Mapping Stat5b interactions in the less stringent data set and more stringent data set

#Defining a function to get 1. Proteins_connected_to_Stat5b, 2. Network plot, 3. Link to webpage, 4. Pathways
subnetworks <- function(data_h,score_h) {
  #a = The data we want to have a network from, b = the threshold number of the string_db function
  string_db <- STRINGdb$new(version="11.0b", species=10090, score_threshold = score_h, input_directory="") #species = mouse 
  get <- string_db$get_subnetwork(data_h) #builds subnetwork from every entry in STRING_id
  ghi <- get[[Stat5b]][1] ##gives the list entry of "get" that lists all 18 entries that interact with Stat5b
  hg <- unique(ghi$`10090.ENSMUSP00000102981`) #gives the unique interaction (9 out of 18), 
  mat <- as_ids(ghi$`10090.ENSMUSP00000102981`) #function as_ids from the igraph package reads the class("igraph.vs")   and collapses it into a matrix
  Proteins_connected_to_Stat5b <- as.data.frame(unique(mat))
  names(Proteins_connected_to_Stat5b)[1] <- "STRING_id" #Renaming column so that add_protein_description recognizes it
  Proteins_connected_to_Stat5b[(nrow(Proteins_connected_to_Stat5b)+1),] <- Stat5b #Adds Stat5b to map its connections
  Proteins_connected_to_Stat5b <- string_db$add_proteins_description(Proteins_connected_to_Stat5b) #Add´s protein information to STRINGid´s
  
  #Plotting the subnetwork
  string_db$plot_network(Proteins_connected_to_Stat5b$STRING_id)
  
  #providing my link
  link <- string_db$get_link(Proteins_connected_to_Stat5b$STRING_id) #Provides link to the STRINGdb page
  
  #Providing the pathways
  pathways <- string_db$get_enrichment(Proteins_connected_to_Stat5b, category = "KEGG")

  return(list(Proteins_connected_to_Stat5b, pathways, link))
  
  remove(get,ghi,hg,mat)
   }

## Less stringent parameter to determine the Proteins that interact with Stat5
subnetwork_less_stringent <- subnetworks(data_h = hits_low$STRING_id, score_h = 1) #Using function
#Sub-setting the returned list into original dataframes + link
Proteins_connected_to_Stat5b_less <- subnetwork_less_stringent[[1]]
pathways_less <- subnetwork_less_stringent[[2]]
Link_less <- subnetwork_less_stringent[[3]]
Link_less #Link to website
invisible(string_db_low$get_png(Proteins_connected_to_Stat5b_less$STRING_id, file = "~/phosphoproteomics_project/results/Network_Stat5_less.png")) #Saving png

## More stringent parameter to determine the Proteins that interact with Stat5
subnetwork_more_stringent <- subnetworks(data_h = hits_high$STRING_id, score_h = 200) #Using function
#Sub-setting the returned list into original dataframes + link
Proteins_connected_to_Stat5b_more <- subnetwork_more_stringent[[1]] 
pathways_more <- subnetwork_more_stringent[[2]]
Link_more <- subnetwork_more_stringent[[3]]
Link_more #Link to website 
invisible(string_db_high$get_png(Proteins_connected_to_Stat5b_more$STRING_id, file = "~/phosphoproteomics_project/results/Network_Stat5_more.png")) #Saving png


```

------------------------------------------------------------------------

#### Printing Pathways

```{r paged.print=TRUE}
#Identifying only for Proteins connected to Stat5b
#Table for less stringent parameters
print.data.frame(pathways_less[1:5,c(3,4,10)], row.names = F)
#Table for more stringent parameters
print.data.frame(pathways_more[1:5,c(3,4,10)], row.names = F)
#Identifying pathways also for Proteins with a ratio >=2 and p.value <= 0.05 and threshold = 200.
enrichment <- string_db_high$get_enrichment(hits_high$STRING_id, category = "KEGG")
print.data.frame(enrichment[1:5,c(3,4,10)], row.names = F)
#Identifying pathways also for Proteins with a ratio >=2 and p.value <= 0.05 and threshold = 1.
enrichment <- string_db_low$get_enrichment(hits_low$STRING_id, category = "KEGG")
print.data.frame(enrichment[1:5,c(3,4,10)], row.names = F)

remove(enrichment,i, Link_less, Link_more, string_db_high, string_db_low, subnetworks, volcano_plot_proteome, subnetwork_less_stringent, subnetwork_more_stringent, hits_high, hits_low, data_phospho)
```

------------------------------------------------------------------------
